Compositional Quantitative Resource Analysis of C Programs

### 2014-06-18 paper/main.tex
small fixes in semantics/logic
more work on section 2

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
notations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
To understand how the analysis works, it is instructive to look at some
more involved examples.

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.22\linewidth}
  \centering
  \begin{minipage}[b]{.20\linewidth}
    \begin{center}
   \begin{lstlisting}
while (n>x) {
  ${\color{blue}%
  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
  if (m>y) 
    ${\color{blue}%
    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
    y = y+1;
    ${\color{blue}%
    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  else
    ${\color{blue}%
    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
    x = x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
}
   \end{lstlisting}

@@

$|[x, n]| + |[z, n]|$
\\[.7\baselineskip]
      {\bf speed\_2}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  while (z-y>0) {
    y = y+1;
    tick(3);
  }
  while (y>9) {
    y=y-10;
    tick(1);
  }
   \end{lstlisting}

$3.1|[y,z]| + 0.1|[0,y]|$
\\[.7\baselineskip] 
      {\bf t08}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  n=n+1;
  y=y+1000;
  while (y>=100 && *){
    y=y-100;
    tick(5);
  }
  tick(9);
}
   \end{lstlisting}

$59|[n,0]| + 0.05|[0,y]|$
\\[.7\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}

  \caption{Derivations of bounds on the number of ticks for
    challenging examples.  In the potential functions, we only mention
    the non-zero terms and in the logical context $\Gamma$ we only
    mention assertions that we actually use in the reasoning.}
  \label{fig:ex_list}
\end{figure*}

@@

For the sake of simplicity, our quantitative program logic
applies to the Clight language.  This language is the most
abstract intermediate language used by CompCert.  It is very
similar to vanilla C but the expressions are free of
side-effects and there is only one looping construct.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~E
\mid x \gets E
\mid x \gets f(E^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\end{align*}
%
We added the builtin primitive $\code{assert}$ that checks
if a condition is true before continuing the program
execution.  This will be used to explicit invariants to the
automatic analysis later.

@@

\and \Rule{S:Call}
{ \delta~f = (x_1, \dots, x_n, S) \\
  ((\state, x_i \mapsto \llbracket e_i \rrbracket_\state), S)
    \Downarrow_{(q,q')} \code{ret}~\state'~v
}
{ (\state, x \gets f(e_1, \dots, e_n)) \Downarrow^r_{M_f\cdot(q,q')}
  \code{seq}~(\state'~\backslash~(x_i))[x \mapsto v]
}

@@

\and
\Rule{L:Call}
{
  \Delta f = \forall z.(P_f\,z, Q_f\,z) \\
  P \models P_f \, y \, (\sem{e_1}_\state,\dots,\sem{e_n}_\state) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \sigma[x \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { x \gets f(e_1, \dots, e_n) }
  { Q }
}

@@

\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z.(P_f\,z, Q_f\,z)
  \rightarrow
  \forall y.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y }{ \delta f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert n>0;
  assert m>0;
  va = n; vb = 0;
  while (va>0 && *) {
    if (vb<m) { 
      vb=vb+1; 
      va=va-1;
    } else {
      vb=vb-1;
      vb=0;
    }
  }
   \end{lstlisting}
$2 + 4|[0, n]|$
\\[.7\baselineskip]
      {\bf fig\_4\_2}
    \end{center}
  \end{minipage}
%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0<m);
  i = n;
  while (i>0 && *) {
    if (i<m)
      i=i-1;
    else
      i=i-m;
  }
   \end{lstlisting}
$|[0, n]|$
\\[.7\baselineskip]
      {\bf fig\_4\_4}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0 < m < n);
  i=m;
  while (0<i<n) {
    if (dir==fwd) i++;
    else i--;
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf fig\_4\_5}
    \end{center}
  \end{minipage}

@@

Our analyzer can derive almost all linear bounds for programs that
have been described as challenges in the literature on bound
generation.  We found only three programs with a linear bound for
which our analyzer could not find a tight bound.  The first one
(\emph{fig\_4\_5}) from [GulwaniPLDI09] requires \emph{path-sensitive
  reasoning} to derive a bound.  The other two, \emph{ex2} and
\emph{ex5} (see [GulwaniPLDI10]), from [GulwaniPLDI10] were described
to have a linear bound but seem to be non-terminating.

@@

Example \emph{fig\_4\_2} from [GulwaniPLDI09] is quite involved.
Amortized reasoning helps to understand how we derive the bound $4 +
4|[0, n]|$.  We start with potential $6 + 4|[0, n]|$ and use the
constant potential to pay for the two first assignments.  We use the
remaining potential $4 + 4|[0, n]|$ and the fact that $vb=0$ to
establish the potential $4 + 4|[0, n]| + 2|[0,vb]|$ that serves as a
loop invariant.  In the \emph{if branch} of the conditional, we use
the constant potential $4$ of the invariant to pay for the two
assignments ($2$ units) and the potential of $|[0,vb]|$ ($2$ units).
Since we also know that $|[0, n]|>0$ we obtain constant potential $4$
and establish the loop invariant again.  In the \emph{else} branch, we
use the potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$
potential units to pay for the assignment.  Since we have $4$ constant
potential units left after the loop, we can subtract them from the
initial potential $6 + 4|[0, n]|$.

In example \emph{fig\_4\_4} it is essential that $m$ is positive.
That ensures that we can obtain constant potential for the interval
size $|[0,i]|$ in the \emph{else} branch of the conditional since
$|[0,i]|$ decreases.  Example \emph{fig\_4\_5} is one of the three
examples that we can not handle automatically.  The execution is
bounded because the boolean value of the test \code{dir==fwd} does not
change during the iteration of the loop.  As a result, the variable
$i$ is either counted down to $0$ or up to $n$.  Our tool cannot
handle example \emph{fig\_4\_5} because we don't do path sensitive
reasoning.  Note however that it would be more efficient to move the
test \code{dir==fwd} outside of the loop (this would be also done by
an optimizing compiler).  The resulting program could be analyzer by our
tool.

>>>>

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
notations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
To understand how the analysis works, it is instructive to look at
some more involved examples.  Examples \emph{speed\_1} and
\emph{speed\_2} are taken from Gulwani et al~\cite{Gulwani-speed}.
They are both handled by the SPEED tool~\cite{Gulwani-speed} but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.22\linewidth}
  \centering
\hspace{-0.4cm}
  \begin{minipage}[b]{.18\linewidth}
    \begin{center}
   \begin{lstlisting}[]
while (n>x) {
  ${\color{blue}%
  n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
  if (m>y) 
    ${\color{blue}%
    m{>}y; \, |[x, n]| {+} |[y, m]| \vdash }$
    y = y+1;
    ${\color{blue}%
    \dashv \cdot ; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  else
    ${\color{blue}%
    n{>}x; \, |[x, n]| {+} |[y, m]| \vdash }$
    x = x+1;
    ${\color{blue}%
    \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| }$
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} |[x, n]| {+} |[y, m]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, |[x, n]| {+} |[y, m]| }$
}
   \end{lstlisting}

@@

$|[x, n]| + |[z, n]|$
\\[.7\baselineskip]
      {\bf speed\_2}
    \end{center}
  \end{minipage}
%
\hfill
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (z-y>0) {
  ${\color{blue}%
  y{<}z; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y = y+1;
  ${\color{blue}%
  \dashv \cdot; \, 3 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(3);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
${\color{blue}%
\dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
while (y>9) {
  ${\color{blue}%
  y{>}9; \, 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  y=y-10;
  ${\color{blue}%
  \dashv \cdot; \, 1 {+} 3.1|[y,z]| {+} 0.1|[0,y]| \vdash }$
  tick(1);
  ${\color{blue}%
  \dashv \cdot; \, 3.1|[y,z]| {+} 0.1|[0,y]| }$
}
   \end{lstlisting}

$3.1|[y,z]| + 0.1|[0,y]|$
\\[.7\baselineskip] 
      {\bf t08}
    \end{center}
  \end{minipage}
%
\hfill
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
while (n<0) {
  ${\color{blue}%
  n{<}0; \,  P(n,y) \vdash }$
  n=n+1;
  ${\color{blue}%
  \dashv \cdot; \,  59 {+} P(n,y) \vdash }$
  y=y+1000;
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  while (y>=100 && *){
    ${\color{blue}%
    y{>}99; \, 9 {+} P(n,y) \vdash }$
    y=y-100;
    ${\color{blue}%
    \dashv y{>}99; \, 14 {+} P(n,y) \vdash }$
    tick(5);
    ${\color{blue}%
    \dashv \cdot; \, 9 {+} P(n,y) }$
  }
  ${\color{blue}%
  \dashv \cdot; \, 9 {+} P(n,y) \vdash }$
  tick(9);
  ${\color{blue}%
  \dashv \cdot; \, P(n,y) \vdash }$
}
   \end{lstlisting}
$59|[n,0]| {+} 0.05|[0,y]|$
\\[.7\baselineskip]
      {\bf t27}
    \end{center}
  \end{minipage}

  \caption{Derivations of bounds on the number of ticks for
    challenging examples.  In the potential functions, we only mention
    the non-zero terms and in the logical context $\Gamma$ we only
    mention assertions that we actually use in the reasoning. In
    Example \emph{t27}, we use the abbreviation $P(n,y) := 59|[n,0]|
    {+} 0.05|[0,y]|$.}
  \label{fig:ex_list}
\end{figure*}

\begin{figure}[t]
  \centering
    \begin{minipage}[b]{\linewidth}
    \begin{center}
   \begin{lstlisting}
void count_down (int x,int y)
{ int a = x;
  if (a>y) {a=a-1; tick(1); count_up(a,y); }
}
void count_up (int x, int y)
{ int a = y;
  if (a+1<x) {a = a+2; tick(1); count_down(x,a); }
}
   \end{lstlisting}

$0.33 + 0.67 |[y,x]|\;\;$ (\code{count\_down(x,y)})\\
$\;\;\;\;\;\;\,0.67 |[y,x]|\;\;$ (\code{count\_up(x,y)})
\\[.7\baselineskip]
      {\bf t39}
    \end{center}
  \end{minipage}
  \caption{Two mutually-recursive functions with the computed tick bounds.}
  \label{fig:ex_rec}
\end{figure}

\subsection{Implementation}

We do not yet generate certificates in the program logic that are
checkable in Coq.  This is substantial additional engineering effort
but is conceptually straightforward as we describe in detail in
Section XXX.

@@

For the sake of simplicity, our quantitative program logic
applies to the Clight language.  This language is the most
abstract intermediate language used by CompCert.  It is very
similar to vanilla C but the expressions are free of
side-effects and there is only one looping construct.
%
\begin{align*}
S &:= \code{assert}~E
\mid \code{skip}
\mid \code{break}
\mid \code{return}~E
\mid x \gets E
\mid x \gets f(x^*)
\\
& \mid \code{loop}~S
\mid \code{if}(E)~S~\code{else}~S
\mid S;S
\end{align*}
%
We added the builtin primitive $\code{assert}$ that checks
if a condition is true before continuing the program
execution.  This will be used to explicit invariants to the
automatic analysis later.

@@

\and \Rule{S:Call}
{ \delta~f = (\vec y, S) \\
  ((\state, y_i \mapsto \state(x_i)), S)
    \Downarrow_{(q,q')} \code{ret}~\state'~v
}
{ (\state, r \gets f(\vec x)) \Downarrow_{M_f\cdot(q,q')}
  \code{seq}~(\state'~\backslash~(x_i))[r \mapsto v]
}

@@

\and
\Rule{L:Call}
{
  \Delta f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z) \\
  P \models P_f \, y \, (\sigma(\vec x)) \land A \\
  \forall v.\, (Q_f \, y \, v \land A \models \lambda \state.\, Q \, \sigma[r \mapsto v])
}
{ \Delta; B; R \vdash
\htriple
  { M_f + P }
  { r \gets f(\vec x) }
  { Q }
}

@@

\and
\Rule{L:Extend}
{ \forall f\,P_f\,Q_f.\,
  \Delta'f = \forall z\,\vec v.(P_f\,z\,\vec v, Q_f\,z)
  \rightarrow
  \forall y\,\vec v.\,
  (\Delta \cup \Delta'; \bot; Q_f\,y \vdash
    \htriple{ P_f\,y\,\vec v }{ \delta f }{ \bot })
\\
  \Delta \cup \Delta'; B; R \vdash
  \htriple{ P }{ S }{ Q }
}
{ \Delta; B; R \vdash
\htriple{ P }{ S }{ Q }
}

@@

\begin{figure*}[t!]
 \setlength{\progwidth}{.24\linewidth}
  \centering
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert n>0;
  assert m>0;
  va = n; vb = 0;
  while (va>0 && *) {
    if (vb<m) { 
      vb=vb+1; 
      va=va-1;
    } else {
      vb=vb-1;
      vb=0;
    }
  }
   \end{lstlisting}
$2 + 4|[0, n]|$
\\[.7\baselineskip]
      {\bf fig4\_2}
    \end{center}
  \end{minipage}
%
%
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0<m);
  i = n;
  while (i>0 && *) {
    if (i<m)
      i=i-1;
    else
      i=i-m;
  }
   \end{lstlisting}
$|[0, n]|$
\\[.7\baselineskip]
      {\bf fig4\_4}
    \end{center}
  \end{minipage}
%
%
  \begin{minipage}[b]{\progwidth}
    \begin{center}
   \begin{lstlisting}
  assert (0 < m < n);
  i=m;
  while (0<i<n) {
    if (dir==fwd) i++;
    else i--;
  }
   \end{lstlisting}
$---$
\\[.7\baselineskip]
      {\bf fig4\_5}
    \end{center}
  \end{minipage}

@@

Our analyzer can derive almost all linear bounds for programs that
have been described as challenges in the literature on bound
generation.  We found only three programs with a linear bound for
which our analyzer could not find a tight bound.  The first one
(\emph{fig4\_5}) from [GulwaniPLDI09] requires \emph{path-sensitive
  reasoning} to derive a bound.  The other two, \emph{ex2} and
\emph{ex5} (see [GulwaniPLDI10]), from [GulwaniPLDI10] were described
to have a linear bound but seem to be non-terminating.

@@

Example \emph{fig4\_2} from [GulwaniPLDI09] is quite involved.
Amortized reasoning helps to understand how we derive the bound $4 +
4|[0, n]|$.  We start with potential $6 + 4|[0, n]|$ and use the
constant potential to pay for the two first assignments.  We use the
remaining potential $4 + 4|[0, n]|$ and the fact that $vb=0$ to
establish the potential $4 + 4|[0, n]| + 2|[0,vb]|$ that serves as a
loop invariant.  In the \emph{if branch} of the conditional, we use
the constant potential $4$ of the invariant to pay for the two
assignments ($2$ units) and the potential of $|[0,vb]|$ ($2$ units).
Since we also know that $|[0, n]|>0$ we obtain constant potential $4$
and establish the loop invariant again.  In the \emph{else} branch, we
use the potential $2|[0,vb]|$ and the fact $vb>0$ to obtain $2$
potential units to pay for the assignment.  Since we have $4$ constant
potential units left after the loop, we can subtract them from the
initial potential $6 + 4|[0, n]|$.

In example \emph{fig4\_4} it is essential that $m$ is positive.
That ensures that we can obtain constant potential for the interval
size $|[0,i]|$ in the \emph{else} branch of the conditional since
$|[0,i]|$ decreases.  Example \emph{fig4\_5} is one of the three
examples that we can not handle automatically.  The execution is
bounded because the boolean value of the test \code{dir==fwd} does not
change during the iteration of the loop.  As a result, the variable
$i$ is either counted down to $0$ or up to $n$.  Our tool cannot
handle example \emph{fig4\_5} because we don't do path sensitive
reasoning.  Note however that it would be more efficient to move the
test \code{dir==fwd} outside of the loop (this would be also done by
an optimizing compiler).  The resulting program could be analyzer by our
tool.
